---
title: "Luigi's Data June 2022"
output: html_notebook
---

On Tue, Jul 12, 2022 at 1:51 PM Luigi Brunetti <luigi.brunetti@rutgers.edu> wrote:
Here are the drugs that could be interesting to look at:

 

Remdesvir

Corticosteroids (just combine dexamethasone, methylprednisone, prednisone, and hydrocortisone)

ACEi

ARBs

Metformin

Sulfonylureas (combine glimepiride and glipizide)

Sitagliptin

Anticoagulation (subgroups therapeutic and prophylactic)

 

If we need to reduce the number then I would remove remdesivir, corticosteroids, and anticoagulation since these would be acute medications versus the others which would be chronic medications.  Let me know if this makes sense.

 

I also asked the lab for more samples, will see what they say.

# Setup
```{r}
require(data.table)
require(ggplot2)
require(table1)
require(randomForest)
require(randomForestExplainer)
require(rpart)
```

# Load data
```{r}
dt1 <- fread("COVID_19_RNA_Demo.csv")

# Rename
colnames(dt1)[colnames(dt1) == "Rheum/ CONNECTIVE TISSUE DISEASE"] <- "Rheum"
colnames(dt1)[colnames(dt1) == "Smoking History (0,1,2)"] <- "Smoking"
```

# Convert variables
```{r}
which(colnames(dt1) == "If BMI ?30 Obesity Class (I,II,III)")
which(colnames(dt1) == "Prophylactic Anticoagulation")
which(colnames(dt1) == "Hyperlipidemia")

dt1[, c(1:3,
        5:6,
        10:15,
        18:61,
        77:85)] <- lapply(X = dt1[, c(1:3,
                                      5:6,
                                      10:15,
                                      18:61,
                                      77:85)],
                          FUN = "factor")

which(colnames(dt1) == "ALT")

dt1[, c(4,
        7:9,
        64:76)] <- lapply(X = dt1[, c(4,
                                      7:9,
                                      64:76)],
                          FUN = "as.numeric")
```

# Response: Critical Care
## Table 1.1
```{r}
table1( ~.|`Critical Care`,
       data = dt1[, -c("Fake_ID")])
```
  
## Chi-sq test
```{r}
chisq.test(dt1$`Critical Care`,
           dt1$Vaccinated,
           correct = FALSE)

chisq.test(dt1$`Critical Care`,
           dt1$Vaccinated,
           correct = TRUE)

chisq.test(dt1$`Critical Care`,
           dt1$Vaccinated,
           simulate.p.value = TRUE)
```

## Logistic regression
```{r}
m1 <- glm(`Critical Care` ~ Vaccinated,
          data = dt1,
          family = "binomial")
summary(m1)

m2 <- glm(`Critical Care` ~ Vaccinated + 
            Sex +
            Race +
            Wave + 
            Age_Cat +
            MI +
            HF +
            PVD +
            CVD +
            Dementia +
            COPD +
            Rheum +
            AnyDM +
            Smoking +
            # PUD +
            # mliverd +
            # Hemoplegia +
            RenalD +
            Cancer +
            HTN +
            Obese +
            Hyperlipidemia,
          data = dt1,
          family = "binomial")
summary(m2)
```

## Poisson regression
```{r}
m3 <- glm((`Critical Care` == 1) ~ Vaccinated,
          data = dt1,
          family = "poisson")
summary(m3)

m4 <- glm((`Critical Care` == 1) ~ Vaccinated + 
            Sex +
            Race +
            Wave + 
            Age_Cat +
            MI +
            HF +
            PVD +
            CVD +
            Dementia +
            COPD +
            Rheum +
            AnyDM +
            Smoking +
            # PUD +
            # mliverd +
            # Hemoplegia +
            RenalD +
            Cancer +
            HTN +
            Obese +
            Hyperlipidemia,
          data = dt1,
          family = "poisson")
summary(m4)
```

## Random forest
```{r}
rf1 <- randomForest(`Critical Care` ~ Vaccinated + 
                      Sex +
                      Race +
                      Wave + 
                      Age_Cat +
                      MI +
                      HF +
                      PVD +
                      CVD +
                      Dementia +
                      COPD +
                      Rheum +
                      AnyDM +
                      Smoking +
                      # PUD +
                      # mliverd +
                      # Hemoplegia +
                      RenalD +
                      Cancer +
                      HTN +
                      Obese +
                      Hyperlipidemia,
                    data = dt1[!is.na(Smoking), ],
                    ntree = 10000,
                    importance = TRUE,
                    localImp = TRUE)
rf1
plot(rf1)
```

```{r}
explain_forest(forest = rf1)
```

## Desision tree
```{r}
rpart1 <- rpart(`Critical Care` ~ Vaccinated + 
                  Sex +
                  Race +
                  Wave + 
                  Age_Cat +
                  MI +
                  HF +
                  PVD +
                  CVD +
                  Dementia +
                  COPD +
                  Rheum +
                  AnyDM +
                  Smoking +
                  # PUD +
                  # mliverd +
                  # Hemoplegia +
                  RenalD +
                  Cancer +
                  HTN +
                  Obese +
                  Hyperlipidemia,
                data = dt1,
                method = "class",
                control = rpart.control(minsplit = 1,
                                        minbucket = 1,
                                        cp = 0))
rpart1
```

```{r}
plot(rpart1)
text(rpart1,
     use.n = TRUE,
     all = TRUE,
     cex = 0.6)

```

# Response: Death
## Table1.2
```{r}
table1( ~.|Expired,
       data = dt1[, -c("Fake_ID")])
```

## Chi-sq test
```{r}
addmargins(table(Expired = dt1$Expired,
                 Vaccinated = dt1$Vaccinated))
100*prop.table(addmargins(table(Expired = dt1$Expired,
                                Vaccinated = dt1$Vaccinated)))


chisq.test(dt1$Expired,
           dt1$Vaccinated,
           correct = TRUE)
```

## Logistic regression
```{r}
m1 <- glm(Expired ~ Vaccinated,
          data = dt1,
          family = "binomial")
summary(m1)

m2 <- glm(Expired ~ Vaccinated + 
            Sex +
            Race +
            Wave + 
            Age_Cat +
            MI +
            HF +
            PVD +
            CVD +
            Dementia +
            COPD +
            Rheum +
            AnyDM +
            Smoking +
            # PUD +
            # mliverd +
            # Hemoplegia +
            RenalD +
            Cancer +
            HTN +
            Obese +
            Hyperlipidemia,
          data = dt1,
          family = "binomial")
summary(m2)
```

## Random forest
```{r}
rf2 <- randomForest(Expired ~ Vaccinated + 
                      Sex +
                      Race +
                      Wave + 
                      Age_Cat +
                      MI +
                      HF +
                      PVD +
                      CVD +
                      Dementia +
                      COPD +
                      Rheum +
                      AnyDM +
                      Smoking +
                      # PUD +
                      # mliverd +
                      # Hemoplegia +
                      RenalD +
                      Cancer +
                      HTN +
                      Obese +
                      Hyperlipidemia,
                    data = dt1[!is.na(Smoking), ],
                    ntree = 10000,
                    importance = TRUE,
                    localImp = TRUE)
rf2
plot(rf2)
```

```{r}
explain_forest(forest = rf2)
```

## Additional tables
```{r}
addmargins(table(dt1$`Critical Care`,
                 dt1$Expired))

chisq.test(dt1$`Critical Care`,
                 dt1$Expired)
```


# Response: log[LOS]
## Table1.3
```{r, warning=FALSE}
table1( ~.|(LOS > 5),
       data = dt1[, -c("Fake_ID")])
```

## Figures
```{r}
hist(dt1$LOS, 110)
table(dt1$Vaccinated)


dt1$logLOS <- log(dt1$LOS + 1)
hist(dt1$logLOS, 50)

boxplot(logLOS ~ Vaccinated,
         data = dt1)
```

## Linear regression
```{r}
m1 <- lm(logLOS ~ Vaccinated,
         data = dt1)
summary(m1)

m2 <- lm(logLOS ~ Vaccinated + 
           Sex +
           Race +
           Wave + 
           Age_Cat +
           MI +
           HF +
           PVD +
           CVD +
           Dementia +
           COPD +
           Rheum +
           AnyDM +
           Smoking +
           # PUD +
           # mliverd +
           # Hemoplegia +
           RenalD +
           Cancer +
           HTN +
           Obese +
           Hyperlipidemia,
         data = dt1)
summary(m2)

plot(dt1$logLOS ~ dt1$BMI)
```

## Random forest
```{r}
rf3 <- randomForest(logLOS ~ Vaccinated + 
                      Sex +
                      Race +
                      Wave + 
                      Age_Cat +
                      MI +
                      HF +
                      PVD +
                      CVD +
                      Dementia +
                      COPD +
                      Rheum +
                      AnyDM +
                      Smoking +
                      # PUD +
                      # mliverd +
                      # Hemoplegia +
                      RenalD +
                      Cancer +
                      HTN +
                      Obese +
                      Hyperlipidemia,
                    data = dt1[!is.na(Smoking), ],
                    ntree = 10000,
                    importance = TRUE,
                    localImp = TRUE)
rf3
plot(rf3)
```

```{r}
explain_forest(forest = rf3)
```

# ELISA data
```{r}
dt21 <- fread("Luigi_August2023/COVID study Master sheet w ELISA data July 2023.csv")
dt22 <- fread("Luigi_August2023/COVID study Master sheet w ELISA data July 2023-2.csv")
```

# Gene expressions
```{r}
dt3 <- fread("Luigi_August2023/gene_counts.txt")
```

# Session
```{r}
sessionInfo()
```

