---
title: "Luigi's COVID-19 Data"
output: html_notebook
---

Version: 3.0  
Created: 9/6/2023  
Last update: 11/16/2023

Analysis datasets and copy of the Rmd are in Rutgers box:  
https://rutgers.app.box.com/folder/227391194192  

GitHub Repo:  
https://github.com/sargdavid/luigi_covid19  
  
# Setup
```{r setup}
options(scipen=999)

require(data.table)
require(ggplot2)
require(table1)
require(DESeq2)

require(tidytlg)
require(gt)
require(rtables)
require(gridExtra)
```

Javier, 9/15/2023:  
1. Build survival model using LOS as dates. Assuming all death are in-hospital, end of LOS.  

Meeting 9/19/23:  
1. Was everybody admitted with COVID-19 symptoms? In Marshall's email from 11/14/23 8:01AM he said "not sure [admission DX] will be necessary since admitting DX will be COVID". Does this mean that controls where also admitted with COVID diagnoses? How?  
2. Do survival for in-patients only using LOS.  
3. Look at LOS of dead patients.  
4. Do DM patients have higher WHO OSCI score at admission?  
5. COVID+DM = more likely to go to ICU vs COVID+noDM?
6. Outcomes: death, LOS, ventilation, ICU.  

Meeting 9/7/23:  
1. Look at inflammation signature w/wo DM, stratified by COVID.  
2. in COVID, coagulation cascade pathway was affected. See if we can find it in clin. data, ELIZA or RNA-seq. SHould coagulation be considered in COVID pateints' treatment?  
3. Coding/non-coding gene co-expression.  
4. Activation of inflammatory pathways because we have transcriptome, not just genome.  
5. Neanderthals' genes (see Lu's email).  
6. Important factors: Critical care, WHO OSCI, LOS, ACE inhibitors, age>55, HTN, DM.  
  
```{r}
# rows_empty <- function(mat) {
#   unlist(apply(mat, 1, function(x) !any(nzchar(x))), simplify = FALSE)
# }
# 
# #' Emit a table as RTF via gentlg
# #' @inheritParams formatters::matrix_form
# tt_to_tbldf <- function(tt) {
#   mpf <- matrix_form(tt, indent_rownames = FALSE, expand_newlines = FALSE)
#   strmat <- mf_strings(mpf)
#   nhl <- mf_nlheader(mpf)
#   strbody <- strmat[-(1:nhl), , drop = FALSE]
#   row_type <- apply(strbody, 1, 
#                     function(x) {
#                       nonempty <- nzchar(x)
#                       if(all(!nonempty)) { ## shouldn't ever happen
#                         "EMPTY"
#                       } else if (nonempty[1] && all(!nonempty[-1])) {# only 1st cell nonempty
#                         "HEADER"
#                       } else {
#                         if(x[1] == "N")
#                           "N"
#                         else
#                           "VALUE"
#                       }
#                     })
#   rowdf <- mf_rinfo(mpf)
#   rinds <- mf_lgrouping(mpf)[-(1:nhl)] - mf_nrheader(mpf)
#   indentme <- rowdf$indent[rinds]
#   anbr <- cumsum(!is.na(c(0, head(rowdf$trailing_sep, -1))))[rinds] + 1 ## so it starts at 1
#   roworder <- 1:NROW(rowdf) - (anbr - 1)
#   newrows <- c(0, ifelse(tail(anbr, -1) == head(anbr, -1), 0, 1))
#   
#   tbldf <- as.data.frame(strbody)
#   names(tbldf) <- c("label", paste("col", seq(1, ncol(tbldf) - 1)))
#   cbind(tbldf, data.frame(row_type = row_type,
#                           anbr = anbr,
#                           indentme = indentme,
#                           roworder = roworder, 
#                           newrows = newrows,
#                           stringsAsFactors = FALSE))
# }
# 
# #' @rdname tt_to_tbldf
# #' @param file character(1). File path for output.
# tt_to_tlgrtf <- function(tt, file = "./table") {
#   df <- tt_to_tbldf(tt)
#   mpf <- matrix_form(tt, indent_rownames = FALSE, expand_newlines = FALSE)
#   
#   strmat <- mf_strings(mpf)
#   nhl <- mf_nlheader(mpf)
#   nspancols <- nhl - 1L
#   if(nspancols > 0)
#     csph <- lapply(seq_len(nspancols), function(ii) strmat[ii,])
#   else
#     csph <- NULL
#   gentlg(df,
#          colspan = csph, 
#          file = file,
#          colheader = strmat[nhl,],
#          title = main_title(tt),
#          footers = c(main_footer(tt), prov_footer(tt)))
# }
```

# Data
## Clinical variables and ELISA
```{r}
dt1 <- fread("Data/luigi_covid19_clinical_data.csv")
```

### Format clinical variables
```{r}
dt1$`Real ID` <- as.character(dt1$`Real ID`)

dt1$`Whole Blood` <- factor(dt1$`Whole Blood`,
                            levels = unique(dt1$`Whole Blood`))

dt1$Plasma <- factor(dt1$Plasma)

# COVID = 0/1
dt1$`Study Group` <- factor(dt1$`Study Group`,
                            levels = c("No COVID",
                                       "COVID"))

dt1$`Inpatient=1, ER=2, OP=3` <- factor(dt1$`Inpatient=1, ER=2, OP=3`,
                                        levels = c("Inpatient",
                                                   "ER",
                                                   "OP"))

dt1$Sex <- factor(dt1$Sex,
                  levels = c("Male",
                             "Female"))

dt1$Race <- factor(dt1$Race,
                   levels = c("White Non-Hispanic",
                              "Black",
                              "Indian/Asian",
                              "Other",
                              "Unknown"))

dt1$`If BMI ?30 Obesity Class (I,II,III)` <- factor(dt1$`If BMI ?30 Obesity Class (I,II,III)`,
                                                    levels = c("I",
                                                               "II",
                                                               "III"))

# !!!GET LABELS!!!
dt1$Vaccinated <- factor(dt1$Vaccinated,
                         levels = c(0, 1, 9))

dt1$`Dose #1` <- factor(dt1$`Dose #1`,
                        levels = c(0, 1, 9))

dt1$`Dose #2` <- factor(dt1$`Dose #2`,
                        levels = c(0, 1, 9))

dt1$Booster <- factor(dt1$Booster,
                      levels = c(0, 1, 9))

dt1$`Critical Care` <- factor(dt1$`Critical Care`,
                              levels = 0:1)

# `ICU LOS` = length of stay in the hospital
dt1$Still_In_Hospital_08182023 <- is.na(as.numeric(dt1$`ICU LOS`))
dt1$`ICU LOS` <- as.numeric(dt1$`ICU LOS`)

# Death
dt1$Expired <- factor(dt1$Expired,
                      levels = 0:1)

dt1$Wave <- factor(dt1$Wave,
                   levels = c(1:4))

# Luigi's keys
dt1$`Smoking History (0,1,2)` <- factor(dt1$`Smoking History (0,1,2)`,
                                        levels = c("Current",
                                                   "Former",
                                                   "Never"))

# !!! GET LABELS!!!
dt1$Age_Category <- factor(dt1$Age_Category,
                               levels = 0:4)

# Co-diagnoses/pre-existing conditions
dt1$MI <- factor(dt1$MI)
dt1$HF <- factor(dt1$HF)
dt1$PVD <- factor(dt1$PVD)
dt1$CVD <- factor(dt1$CVD)
dt1$Dementia <- factor(dt1$Dementia)
dt1$COPD <- factor(dt1$COPD)
dt1$`Rheum/ CONNECTIVE TISSUE DISEASE` <- factor(dt1$`Rheum/ CONNECTIVE TISSUE DISEASE`)
dt1$PUD <- factor(dt1$PUD)
dt1$mliverd <- factor(dt1$mliverd)
dt1$DiabNC <- factor(dt1$DiabNC)
dt1$DiabC <- factor(dt1$DiabNC)
dt1$AnyDM <- factor(dt1$AnyDM)
dt1$Hemoplegia <- factor(dt1$Hemoplegia)
dt1$RenalD <- factor(dt1$RenalD)
dt1$Cancer <- factor(dt1$Cancer)
dt1$msliver <- factor(dt1$msliver)
dt1$MetaCancer <- factor(dt1$MetaCancer)
dt1$AIDS <- factor(dt1$AIDS)
dt1$HTN <- factor(dt1$HTN)
dt1$Obese <- factor(dt1$Obese)

# Medications
dt1$tocilizumab <- factor(dt1$tocilizumab)
dt1$remdesivir <- factor(dt1$remdesivir)
dt1$dexamethasone <- factor(dt1$dexamethasone)
dt1$methylprednisolone <- factor(dt1$methylprednisolone)
dt1$prednisone <- factor(dt1$prednisone)
dt1$hydrocortisone <- factor(dt1$hydrocortisone)
dt1$azithromycin <- factor(dt1$azithromycin)
dt1$hydroxychloroquine <- factor(dt1$hydroxychloroquine)
dt1$regeneron <- factor(dt1$regeneron)
dt1$ACEis <- factor(dt1$ACEis)
dt1$ARBs <- factor(dt1$ARBs)
dt1$ARNIs <- factor(dt1$ARNIs)
dt1$Insulin <- factor(dt1$Insulin)
dt1$Metformin <- factor(dt1$Metformin)
dt1$Glimepiride <- factor(dt1$Glimepiride)
dt1$Glipizide <- factor(dt1$Glipizide)
dt1$Sitagliptin <- factor(dt1$Sitagliptin)
dt1$`Full Dose Anticoagulation` <- factor(dt1$`Full Dose Anticoagulation`)
dt1$`Prophylactic Anticoagulation` <- factor(dt1$`Prophylactic Anticoagulation`)

# Lab work
## NOTE: set correct values for LOQ/LOD! Converted to NAs for now
dt1$HbA1c <- as.numeric(dt1$HbA1c)
dt1$Vit_D_level <- as.numeric(dt1$Vit_D_level)
dt1$ALT <- as.numeric(dt1$ALT)
dt1$AST <- as.numeric(dt1$AST)
dt1$`hs-CRP` <- as.numeric(dt1$`hs-CRP`)
dt1$CRP <- as.numeric(dt1$CRP)
dt1$Fibrinogen <- as.numeric(dt1$Fibrinogen)
dt1$`D-Dimer` <- as.numeric(dt1$`D-Dimer`)
dt1$ESR <- as.numeric(dt1$ESR)
dt1$Ferritin <- as.numeric(dt1$Ferritin)
dt1$LDH <- as.numeric(dt1$LDH)
dt1$`Lactic Acid` <- as.numeric(dt1$`Lactic Acid`)
dt1$PCT <- as.numeric(dt1$PCT)
dt1$`Glucose Lvl` <- as.numeric(dt1$`Glucose Lvl`)
dt1$Screat <- as.numeric(dt1$Screat)

# Diagnoses
dt1$Hyperlipidemia <- factor(dt1$Hyperlipidemia)
dt1$`# Metabolic Syndrome Criteria` <- factor(dt1$`# Metabolic Syndrome Criteria`)

# NA should be non-COVID patients - CHECK
dt1$`WHO-OSCI D1` <- factor(dt1$`WHO-OSCI D1`,
                            levels = 0:7)
dt1$`WHO-OSCI D3` <- factor(dt1$`WHO-OSCI D3`,
                            levels = 0:7)
dt1$`WHO-OSCI D7` <- factor(dt1$`WHO-OSCI D7`,
                            levels = 0:7)
dt1$`WHO-OSCI D14` <- factor(dt1$`WHO-OSCI D14`,
                             levels = 0:7)
dt1$`WHO-OSCI D21` <- factor(dt1$`WHO-OSCI D21`,
                             levels = 0:7)
dt1$`WHO-OSCI D28` <- factor(dt1$`WHO-OSCI D28`,
                             levels = 0:7)

# What is the scale here? 0-7? but we have 8s!
dt1$`WHO-OSCI DISCHARGE` <- factor(dt1$`WHO-OSCI DISCHARGE`)

# ELISA: all numeric
```

### Summary clinical and ELISA
```{r}
summary(dt1)
```

## Gene expressions
NOTE: data contains both, coding and non-coding genes.  
  
### First RNA-seq file
```{r}
dt2 <- fread("Data/gene_count.xls")

rnaseq_meta1 <- data.table(sample_names = colnames(dt2)[2:58],
                           rnaseq_cohort = sapply(strsplit(x = colnames(dt2)[2:58], 
                                                           split = "_"), 
                                                  function(x) x[1]),
                           `Real ID` = sapply(strsplit(x = colnames(dt2)[2:58], 
                                                       split = "_"), 
                                              function(x) x[2]))
rnaseq_meta1
```

### Second RNA-seq file
```{r}
dt3 <- fread("Data/gene_count_Additional.txt")

rnaseq_meta2 <- data.table(sample_names = names(dt3)[2:46],
                           rnaseq_cohort = "",
                           `Real ID` = gsub(pattern = "LB", 
                                            x = names(dt3)[2:46],
                                            replacement = ""))
rnaseq_meta2
```

### Merge RNA-seq meta files
```{r}
rnaseq_meta <- rbindlist(list(rnaseq_meta1,
                              rnaseq_meta2))
length(unique(rnaseq_meta$`Real ID`))
# Same as number of rows, hence, all unique sample IDs

rnaseq_meta$rnaseq_cohort <- NULL
```

### Merge RNA-seq files
First 10 columns are gene info, the other 102 columns are the samples.
```{r}
dt_rna <- merge(dt2,
                dt3,
                by = c("gene_name",
                       "gene_id",
                       "gene_chr",
                       "gene_start",
                       "gene_end",
                       "gene_strand",
                       "gene_length",
                       "gene_biotype",
                       "gene_description",
                       "tf_family"))
```

### IDs that are in RNA-seq but not in Clinical
11 IDs  
```{r}
id_no_clin <- rnaseq_meta$`Real ID`[!(rnaseq_meta$`Real ID` %in% dt1$`Real ID`)]
id_no_clin
```

### Patient "7215440" has 2 RNA-seq samples. We will use the 2nd one "7215440b" and discard the 1st one "7215440a"
```{r}
dt1[`Real ID` == "7215440"]

rnaseq_meta$`Real ID`[rnaseq_meta$`Real ID` == "7215440b"] <- "7215440"
rnaseq_meta$`Real ID`[!(rnaseq_meta$`Real ID` %in% dt1$`Real ID`)]
```

### IDs that are in Clinical but not in RNA-seq
91 IDs  
```{r}
id_no_rna <- dt1$`Real ID`[!(dt1$`Real ID` %in% rnaseq_meta$`Real ID`)]
id_no_rna 
```

### lncRNA database from John Johnson (NOT USED YET)
```{r}
# dt_lnc_john <- fread("Data/lncipedia_transcript_type.lncRNA")
```

## Save data
```{r}
save(dt1,
     file = "Data/dt1.RData")

save(dt_rna,
     file = "Data/dt_rna.RData")

save(rnaseq_meta,
     file = "Data/rnaseq_meta.RData")

write.csv(dt1,
          file = "Data/luigi_covid19_clinical_data.csv",
          row.names = FALSE)

write.csv(dt_rna,
          file = "Data/luigi_covid19_nra_seq_data.csv",
          row.names = FALSE)
```

## Clean
```{r}
rm(list = ls())
gc()
```

# Reload data
```{r}
load("Data/dt1.RData")
load("Data/dt_rna.RData")
load("Data/rnaseq_meta.RData")
```

# Analysis
## Table 1
```{r}
t1 <- table1( ~.| `Study Group` + AnyDM,
              data = dt1[, c(-1)])

t1
```

```{r}
write.table (t1 , 
             file = "Output/table1.csv",
             col.names = TRUE, 
             row.names = FALSE,
             append = TRUE,
             sep = ',')
```

## Death
### Death vs. COVID
```{r}
addmargins(table(Death = dt1$Expired,
                 COVID = dt1$`Study Group`))
```

```{r}
m1 <- glm(Expired ~ `Study Group`,
          family = binomial,
          data = dt1)
s1 <- summary(m1)
ci1 <- confint(m1)

out1 <- data.table(Response = "Dead",
                    Comparison = "COVID vs. No Covid",
                    OR = round(exp(s1$coefficients[-1, 1]),
                               2),
                    `95% C.I.L.B.` = exp(ci1[2, 1]),
                    `95% C.I.U.B.` = exp(ci1[2, 2]),
                    `p-Value` = ifelse(s1$coefficients[-1, 4] >= 0.001,
                                       round(s1$coefficients[-1, 4],
                                             3),
                                       "<0.001"),
                    sign = ifelse(s1$coefficients[-1, 4] <= 0.05,
                                  ifelse(s1$coefficients[-1, 4] <= 0.01,
                                         "**",
                                         "*"), ""))

out1
```

### Death vs DM,  COVID only 
```{r}
covid <- droplevels(dt1[`Study Group` == "COVID", ])

chisq.test(covid$Expired,
           covid$AnyDM)

addmargins(table(Death = covid$Expired,
                 AnyDM = covid$AnyDM))

nocovid <- droplevels(dt1[`Study Group` != "COVID", ])
```

### Death vs Obese, COVID only 
```{r}
chisq.test(covid$Expired,
           covid$Obese)

addmargins(table(Death = covid$Expired,
                 Obese = covid$Obese))
```

### Death vs BMI, COVID only 
```{r}
t.test(covid$BMI ~ covid$Expired)

ggplot(covid,
       aes(x = Expired,
           y = BMI)) +
  geom_boxplot(outlier.shape = "") +
  geom_point(aes(group = `Real ID`),
             shape = 21,
             size = 2,
             position = position_dodge(0.3)) +
  ggtitle("COVID Patients Only") +
  theme_bw()
```

### Death vs Critical Care
Significant
```{r}
addmargins(table(Death = covid$Expired,
                 CrCare = covid$`Critical Care`))

chisq.test(covid$Expired,
           covid$`Critical Care`)

m3 <- glm(Expired ~ `Critical Care`,
          family = binomial,
          data = covid)
s3 <- summary(m3)

out3 <- data.table(Response = "Dead",
                    Comparison = c("ICU: Yes vs. No"),
                    OR = round(exp(s3$coefficients[-1, 1]),
                               2),
                    `95% C.I.L.B.` = exp(s3$coefficients[-1, 1] - 1.96*s3$coefficients[-1, 2]),
                    `95% C.I.U.B.` = exp(s3$coefficients[-1, 1] + 1.96*s3$coefficients[-1, 2]),
                    `p-Value` = ifelse(s3$coefficients[-1, 4] >= 0.001,
                                       round(s3$coefficients[-1, 4], 3),
                                       "<0.001"),
                    sign = ifelse(s3$coefficients[-1, 4] <= 0.05,
                                  ifelse(s3$coefficients[-1, 4] <= 0.01,
                                         "**",
                                         "*"), ""))

out3

m3 <- glm(Expired ~ `Critical Care`,
          family = binomial,
          data = nocovid)
s3 <- summary(m3)

out31 <- data.table(Response = "Dead",
                    Comparison = c("ICU: Yes vs. No"),
                    OR = round(exp(s3$coefficients[-1, 1]),
                               2),
                    `95% C.I.L.B.` = exp(s3$coefficients[-1, 1] - 1.96*s3$coefficients[-1, 2]),
                    `95% C.I.U.B.` = exp(s3$coefficients[-1, 1] + 1.96*s3$coefficients[-1, 2]),
                    `p-Value` = ifelse(s3$coefficients[-1, 4] >= 0.001,
                                       round(s3$coefficients[-1, 4], 3),
                                       "<0.001"),
                    sign = ifelse(s3$coefficients[-1, 4] <= 0.05,
                                  ifelse(s3$coefficients[-1, 4] <= 0.01,
                                         "**",
                                         "*"), ""))

out31
```

### Death vs. COVID | Diabetes
No significant difference.  
Model with interaction is worse.  
  
```{r}
m2 <- glm(Expired ~ `Study Group` +
            AnyDM,
          family = binomial,
          data = dt1)
s2 <- summary(m2)

out2 <- data.table(Response = "Dead",
                    Comparison = c("COVID vs. No Covid",
                                   "Any DM vs. No DM"),
                    OR = round(exp(s2$coefficients[-1, 1]),
                               2),
                    `95% C.I.L.B.` = exp(s2$coefficients[-1, 1] - 1.96*s2$coefficients[-1, 2]),
                    `95% C.I.U.B.` = exp(s2$coefficients[-1, 1] + 1.96*s2$coefficients[-1, 2]),
                    `p-Value` = ifelse(s2$coefficients[-1, 4] >= 0.001,
                                       round(s2$coefficients[-1, 4], 3),
                                       "<0.001"),
                    sign = ifelse(s2$coefficients[-1, 4] <= 0.05,
                                  ifelse(s2$coefficients[-1, 4] <= 0.01,
                                         "**",
                                         "*"), ""))

out2
```

### Death vs COVID, adjusted for critical care
Both significant in additive model.  
```{r}
m4 <- glm(Expired ~ `Study Group` + `Critical Care`,
          family = binomial,
          data = dt1)
s4 <- summary(m4)
ci4 <- confint(m4)

out4 <- data.table(Response = "Dead",
                    Comparison = c("COVID vs. No Covid",
                                   "ICU"),
                    OR = round(exp(s4$coefficients[-1, 1]),
                               2),
                    `95% C.I.L.B.` = exp(ci4[-1, 1]),
                    `95% C.I.U.B.` = exp(ci4[-1, 2]),
                    `p-Value` = ifelse(s4$coefficients[-1, 4] >= 0.001,
                                       round(s4$coefficients[-1, 4],
                                             3),
                                       "<0.001"),
                    sign = ifelse(s4$coefficients[-1, 4] <= 0.05,
                                  ifelse(s4$coefficients[-1, 4] <= 0.01,
                                         "**",
                                         "*"), ""))

out4
```

No significance in interaction model.
```{r}
m41 <- glm(Expired ~ `Study Group`*`Critical Care`,
          family = binomial,
          data = dt1)
summary(m41)
```


### Death vs COVID, adjusted for DM
No significance
```{r}
m5 <- glm(Expired ~ `Study Group` +
            AnyDM,
          family = binomial,
          data = dt1)
summary(m5)
```

### Death vs COVID, adjusted 
```{r}
m6 <- glm(Expired ~ `Study Group` +
            AnyDM +
            Age_Category +
            Sex +
            Race +
            BMI,
          family = binomial,
          data = dt1)
summary(m6)
```

### New variable: DM+Obesity
```{r}
chisq.test(dt1$AnyDM,
           dt1$Obese)
```

## ICU
### ICU LOS
```{r}
summary(dt1$`ICU LOS`)
hist(dt1$`ICU LOS`, 200)
hist(log2(dt1$`ICU LOS` + 1), 200)

m7 <- lm(log2(`ICU LOS` + 1) ~ AnyDM +
           Age +
           Sex +
           Race +
           BMI +
           `Critical Care`,
         family = binomial,
         data = dt1)
summary(m7)
```

### Critical care
```{r}
m8 <- glm(`Critical Care` ~ AnyDM +
            Age +
            Sex +
            Race +
            BMI,
          family = binomial,
          data = droplevels(dt1))
summary(m8)
```

## WHO-OSCI
### Desease severity at baseline by WHO-OSCI
if <5 = mild  
if >= 5 = severe  
  5 = very sick  
  6 = intubated  
  7 = ventilated  
if 8: dead   

```{r}
covid$who_osci_d1_cat <- "Dead"
covid$who_osci_d1_cat[as.numeric(as.character(covid$`WHO-OSCI D1`)) < 8] <- "Severe"
covid$who_osci_d1_cat[as.numeric(as.character(covid$`WHO-OSCI D1`)) < 5] <- "Mild"

covid$who_osci_d1_cat <- factor(covid$who_osci_d1_cat,
                             levels = c("Mild",
                                        "Severe"))
summary(covid$who_osci_d1_cat)
```

```{r}
addmargins(table("WHO-OSCI D1" = dt1$`WHO-OSCI D1`,
                 CC = dt1$`Critical Care`))
```

### WHO-OSCI over time
```{r}
tmp <- dt1[, c("Real ID",
               "Study Group",
               "AnyDM",
               "Expired",
               "WHO-OSCI D1",
               "WHO-OSCI D3",
               "WHO-OSCI D7",
               "WHO-OSCI D14",
               "WHO-OSCI D21",
               "WHO-OSCI D28",
               "WHO-OSCI DISCHARGE")]

tmp <- melt.data.table(tmp,
                       id.vars = c(1:4),
                       variable.name = "Day",
                       value.name = "WHO_OSCI")
tmp$Day <- factor(tmp$Day,
                       levels = unique(tmp$Day))

tmp <- tmp[!is.na(tmp$WHO_OSCI), ]
ggplot(tmp,
       aes(x = Day,
           y = WHO_OSCI,
           group = `Real ID`,
           color = `Real ID`)) +
  facet_wrap(~ AnyDM) +
  geom_line(position = position_dodge(0.3)) +
  theme_bw() +
  theme(legend.position = "none") +
  theme(axis.text.x = element_text(angle = 45,
                                   hjust = 1))
```

```{r}
table(dt1$`WHO-OSCI D1`,
      dt1$Expired)
```

```{r}
t2 <- dcast.data.table(`Real ID` ~ Day,
                 value.var = "WHO_OSCI",
                 data = tmp[Expired == 1, ])

write.csv(t2,
          file = "tmp/table2.csv",
          row.names = FALSE)

t2
```

```{r}
dt1[,
    .(N = .N,
      mu_los = mean(LOS),
      sem_los = sd(LOS)/sqrt(.N)),
    by = c("Study Group")]

dt1[,
    .(N = .N,
      mu_los = mean(LOS),
      sem_los = sd(LOS)/sqrt(.N)),
    by = c("Expired")]

dt1[,
    .(N = .N,
      mu_los = mean(LOS),
      sem_los = sd(LOS)/sqrt(.N)),
    by = c("Expired",
           "Study Group")]

dt1[LOS > 1,
    .(N = .N,
      mu_los = mean(LOS),
      sem_los = sd(LOS)/sqrt(.N)),
    by = c("Expired",
           "Study Group")]
```


```{r}
dt1$LOS[dt1$`Real ID` == 771982]
```

# RNA-seq
https://bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html  
  
## Merge RNA-seq meta-data with clinical
```{r}
rnaseq_meta_clin <- merge(rnaseq_meta,
                          dt1[, c("Real ID",
                                  "Study Group",
                                  "Expired",
                                  "Critical Care",
                                  "Age",
                                  "Sex",
                                  "Race",
                                  "BMI",
                                  "AnyDM")],
                          by = "Real ID")
```

## Duplicated gene names
```{r}
dups <- dt_rna[gene_name %in% gene_name[duplicated(gene_name) == TRUE], ]
```

## Gene types
```{r}
addmargins(table(dt_rna$gene_biotype))
```

## Protein coding genes only
```{r}
dt_rna_cod <- dt_rna[gene_biotype == "protein_coding",]
```

### Duplicated coding genes
```{r}
dup_cod <- dt_rna_cod[gene_name %in% gene_name[duplicated(gene_name) == TRUE], ]
dup_cod_genes <- unique(dup_cod$gene_name)
dup_cod_genes
```

### Add counts for the same genes
```{r}
out <- list()
for (i in 1:length(dup_cod_genes)) {
  out[[i]] <- dup_cod[gene_name == dup_cod_genes[i], ]
  
  for (j in 11:ncol(dup_cod)) {
    out[[i]][, j] <- sum(out[[i]][, j, with = FALSE])
  }
  out[[i]] <- out[[i]][gene_length == max(gene_length), ][1, ]
}
out <- do.call("rbind",
               out)

out
```

### Delete duplicates and replace them with aggregated counts for those genes
```{r}
dt_rna_cod <- dt_rna_cod[!(gene_name %in% dup_cod_genes), ]

dt_rna_cod <- rbindlist(list(dt_rna_cod,
                             out))

length(unique(dt_rna_cod$gene_name))
nrow(dt_rna_cod)
```

### Prepare DESeq object
```{r}
names(rnaseq_meta_clin)[3] <- "Cohort"

ndx <- which(colnames(dt_rna_cod) %in% rnaseq_meta_clin$sample_names)
rnaseq_meta_clin$Cohort <- gsub(x = rnaseq_meta_clin$Cohort,
                                pattern = " ",
                                replacement = "_")
rnaseq_meta_clin$Cohort <- factor(rnaseq_meta_clin$Cohort,
                                  levels = c("No_COVID",
                                             "COVID"))
```

### DESeq: COVID/non-COVID
```{r}
tmp <- as.matrix(dt_rna_cod[, ndx, with = FALSE])
rownames(tmp) <- dt_rna_cod$gene_id

dds_cod <- DESeqDataSetFromMatrix(countData = tmp, 
                              colData = rnaseq_meta_clin,
                              ~ Cohort)
```

#### Filter out genes with small number of hits
Keep the genes with at least 5 out of 92 samples having at least 10 hits each
```{r}
# Number of samples with at least 10 hits in at least X samples
min_n_hits <- c(1, 5, 10, 20, 50)

l2 <- list()

for (j in 1:length(min_n_hits)) {
  
  l1 <- list()
  
  for (i in 1:length(ndx)) {
    
    l1[[i]] <- data.table(`Min N of hits per sample` = min_n_hits[j],
                          `N samples with min hits or more` = i,
                          `N genes` = sum(rowSums(counts(dds_cod) >= j) >= i))
  }
  l2[[j]] <- do.call("rbind",
                     l1)
}

ll <- do.call("rbind",
              l2)
ll$`Min N of hits per sample` <- factor(ll$`Min N of hits per sample`,
                                        levels = min_n_hits)
```

```{r}
ll$lw <- 0.5
ll$lw[ll$`Min N of hits per sample` == 10] <- 1.5

ggplot(ll,
       aes(x = `N samples with min hits or more`,
           y = `N genes`,
           group = `Min N of hits per sample`,
           color = `Min N of hits per sample`)) +
  geom_line(linewidth = ll$lw) +
  geom_vline(xintercept = 10,
             linetype = "dashed",
             color = "red") +
  scale_x_continuous(breaks = seq(0, 92, 5)) +
  theme_bw()
```

```{r}
keep <- rowSums(counts(dds_cod) >= 10) >= 10
sum(keep)

dds_cod <- dds_cod[keep, ]
```

#### Run DESeq
```{r}
deseq_cod <- DESeq(dds_cod)

# deseq <- DESeq(dds,
#                fitType = "glmGamPoi")

resultsNames(deseq_cod)
```

#### DESeq results
```{r}
res <- results(deseq_cod)
res[1:10, ]
rownames(res)[1:10]
```

```{r}
out <- data.table(gene_id = rownames(res),
                  baseMean = res$baseMean,
                  log2FoldChange = res$log2FoldChange,
                  log2FoldChangeSE = res$lfcSE,
                  stat = res$stat,
                  pvalue = res$pvalue,
                  padj = res$padj)

out <- merge(dt_rna_cod[, c("gene_name",
                            "gene_id")],
             out,
             by = "gene_id")

out <- out[!is.na(padj), ]

setorder(out, 
         padj)


head(out)

write.csv(out,
          file = "tmp/deseq2_cod_covid_nocovid.CSV",
          row.names = FALSE)
```

```{r}
tmp5 <- out[abs(log2FoldChange) > 1 &
              padj < 0.05, ]
# 
# tmp5_up <- out[log2FoldChange > 1 &
#               padj < 0.05, ]
# 
# tmp5_down <- out[log2FoldChange < -1 &
#               padj < 0.05, ]
```

#### Calculate the mean expressions of the selected genes
```{r}
tmp <- dt_rna_cod[gene_name %in% tmp5$gene_name, ]
tmp <- melt.data.table(data = tmp,
                       id.vars = 1,
                       measure.vars = 11:ncol(tmp),
                       variable.name = "sample_names",
                       value.name = "Hits")

tmp <- merge(tmp,
             rnaseq_meta_clin,
             by = "sample_names")
```

```{r,fig.width=10,fig.height=10}
p1 <- ggplot(tmp,
             aes(x = `Study Group`,
                 y = log2(Hits + 1))) +
  facet_wrap(~ gene_name,
             scale = "free_y") +
  geom_boxplot(width = 0.3,
               outlier.shape = "") +
  geom_point(aes(fill = AnyDM,
                 group = sample_names),
             size = 2,
             shape = 21,
             alpha = 0.5,
             position = position_dodge(0.3)) +
  scale_y_continuous("Number of hits",
                     breaks = seq(from = 1,
                                  to = 20,
                                  by = 2),
                     labels = (2^seq(from = 1,
                                     to = 20,
                                     by = 2)) - 1) +
  scale_fill_discrete("Any DM",
                      breaks = c(0, 1),
                      labels = c("No",
                                 "Yes")) +
  ggtitle(paste(c("Genes significantly differentially expressed in COVID-19 vs non-COVID patients.",
                  "Difference of the means of log2[hits+1] > 1 and adjusted p-value < 0.05"),
                collapse = "\n"))+
  theme_bw() +
  theme(legend.position = "bottom")

tiff(filename = "tmp/fig1_sign_genes_cod.tiff",
     height = 10,
     width = 10,
     units = 'in',
     res = 600,
     compression = "lzw+p")
print(p1)
graphics.off()

p1
```

```{r,fig.height=5,fig.width=8}
p2 <- ggplot(tmp[gene_name == "IFI27", ],
             aes(x = `Study Group`,
                 y = log2(Hits + 1))) +
  geom_boxplot(width = 0.3,
               outlier.shape = "") +
  geom_point(aes(fill = Expired,
                 group = sample_names),
             size = 3,
             shape = 21,
             alpha = 0.8,
             position = position_dodge(0.3)) +
  scale_y_continuous("Number of hits",
                     breaks = seq(from = 1,
                                  to = 20,
                                  by = 2),
                     labels = (2^seq(from = 1,
                                     to = 20,
                                     by = 2)) - 1) +
  scale_fill_discrete("In-Hospital Death",
                      breaks = c(0, 1),
                      labels = c("No",
                                 "Yes")) +
  ggtitle(paste(c("IFI27 expression in COVID-19 vs non-COVID patients.",
                  "Difference of the means of log2[hits+1] > 1",
                  "and adjusted p-value < 0.05"),
                collapse = "\n")) +
  theme_bw() +
  theme(legend.position = "bottom")

p3 <- ggplot(tmp[gene_name == "IFI27", ],
             aes(x = `Study Group`,
                 y = log2(Hits + 1))) +
  geom_boxplot(width = 0.3,
               outlier.shape = "") +
  geom_point(aes(fill = `Critical Care`,
                 group = sample_names),
             size = 3,
             shape = 21,
             alpha = 0.8,
             position = position_dodge(0.3)) +
  scale_y_continuous("Number of hits",
                     breaks = seq(from = 1,
                                  to = 20,
                                  by = 2),
                     labels = (2^seq(from = 1,
                                     to = 20,
                                     by = 2)) - 1) +
  scale_fill_discrete("Critical Care",
                      breaks = c(0, 1),
                      labels = c("No",
                                 "Yes")) +
  ggtitle(paste(c("IFI27 expression in COVID-19 vs non-COVID patients.",
                  "Difference of the means of log2[hits+1] > 1",
                  "and adjusted p-value < 0.05"),
                collapse = "\n")) +
  theme_bw() +
  theme(legend.position = "bottom")

tiff(filename = "tmp/fig2_ifi27_covid_noncovid.tiff",
     height = 6,
     width = 9,
     units = 'in',
     res = 600,
     compression = "lzw+p")
grid.arrange(p2,
             p3,
             nrow = 1)
# grid.arrange(p2,
#              p3,
#              nrow = 1,
#              top = textGrob(paste(c("IFI27 expression in COVID-19 vs non-COVID patients.",
#                                     "Difference of the means of log2[hits+1] > 1",
#                                     "and adjusted p-value < 0.05"),
#                                   collapse = "\n"),
#                             gp = gpar(fontsize = 20,
#                                       font = 3)))
graphics.off()

grid.arrange(p2,
             p3,
             nrow = 1)
```

#### Table 3: COVID/non-COVID significant DEs
```{r}
# mu <- tmp[, .(mu = exp(mean(log(Hits + 1))) - 1,
#               med = median(Hits)),
#               by = list(gene_name,
#                         Cohort)]
# mu

med <- tmp[, .(med = median(Hits)),
              by = list(gene_name,
                        Cohort)]
med <- dcast.data.table(data = med,
                       gene_name ~ Cohort,
                       value.var = "med")
med

t3 <- data.table(gene_name = tmp5$gene_name,
                 `Log2 Fold Change` = round(tmp5$log2FoldChange, 2),
                 `SE of Log2 Fold Change` = round(tmp5$log2FoldChangeSE, 2),
                 `Adjusted p-Value` = round(tmp5$padj, 3))

t3 <- merge(med,
            t3,
            by = "gene_name")

names(t3)[1:3] <- c("Gene",
               "Median hits in non-COVID",
               "Median hits in COVID")

write.csv(t3,
          file = "tmp/t3.csv",
          row.names = FALSE)
t3
```

### DESeq: AnyDM/no DM
```{r}
tmp <- as.matrix(dt_rna_cod[, ndx, with = FALSE])
rownames(tmp) <- dt_rna_cod$gene_id

dds_cod_dm <- DESeqDataSetFromMatrix(countData = tmp, 
                                     colData = rnaseq_meta_clin,
                                     ~ AnyDM)
```

#### Filter out genes with small number of hits
```{r}
keep <- rowSums(counts(dds_cod_dm) >= 10) >= 10
sum(keep)

dds_cod_dm <- dds_cod_dm[keep, ]
```

#### Run DESeq
```{r}
deseq_cod_dm <- DESeq(dds_cod_dm)

resultsNames(deseq_cod_dm)
```

#### DESeq results
```{r}
res_dm <- results(deseq_cod_dm)
res_dm[1:10, ]
rownames(res_dm)[1:10]
```

```{r}
out_dm <- data.table(gene_id = rownames(res_dm),
                  baseMean = res_dm$baseMean,
                  log2FoldChange = res_dm$log2FoldChange,
                  log2FoldChangeSE = res_dm$lfcSE,
                  stat = res_dm$stat,
                  pvalue = res_dm$pvalue,
                  padj = res_dm$padj)

out_dm <- merge(dt_rna_cod[, c("gene_name",
                            "gene_id")],
             out_dm,
             by = "gene_id")

out_dm <- out_dm[!is.na(padj), ]

setorder(out_dm, 
         padj)


head(out_dm)

write.csv(out_dm,
          file = "tmp/deseq2_cod_dm.CSV",
          row.names = FALSE)
```

```{r}
tmp6 <- out_dm[abs(log2FoldChange) > 1 &
              padj < 0.05, ]
```

#### Calculate the mean expressions of the selected genes
```{r}
tmp <- dt_rna_cod[gene_name %in% tmp6$gene_name, ]
tmp <- melt.data.table(data = tmp,
                       id.vars = 1,
                       measure.vars = 11:ncol(tmp),
                       variable.name = "sample_names",
                       value.name = "Hits")

tmp <- merge(tmp,
             rnaseq_meta_clin,
             by = "sample_names")
```

#### Table 3: DM/no DM significant DEs
```{r}
med <- tmp[, .(med = median(Hits)),
           by = list(gene_name,
                     Cohort)]
med <- dcast.data.table(data = med,
                        gene_name ~ Cohort,
                        value.var = "med")
med

t4 <- data.table(gene_name = tmp6$gene_name,
                 `Log2 Fold Change` = round(tmp6$log2FoldChange, 2),
                 `SE of Log2 Fold Change` = round(tmp6$log2FoldChangeSE, 2),
                 `Adjusted p-Value` = round(tmp6$padj, 3))

t4 <- merge(med,
            t4,
            by = "gene_name")

names(t4)[1:3] <- c("Gene",
                    "Median hits in non-COVID",
                    "Median hits in COVID")

write.csv(t4,
          file = "tmp/t4.csv",
          row.names = FALSE)
t4
```

```{r}
sum(t4$`Log2 Fold Change` > 0)
sum(t4$`Log2 Fold Change` < 0)
```

### DE in COVID and DM
```{r}
covid_and_dm <- t4$Gene[t4$Gene %in% t3$Gene]
covid_and_dm
```

```{r}
dt_f3 <- dt_rna_cod[gene_name %in% covid_and_dm, ]

dt_f3 <- melt.data.table(data = dt_f3,
                         id.vars = 1,
                         measure.vars = 11:ncol(dt_f3),
                         variable.name = "sample_names",
                         value.name = "Hits")
dt_f3 <- merge(dt_f3,
               rnaseq_meta_clin,
               by = "sample_names")
```

```{r,fig.height=10,fig.width=6}
p4 <- ggplot(dt_f3,
             aes(x = Cohort,
                 y = log2(Hits + 1))) +
  facet_wrap(~ gene_name,
             scale = "free_y",
             ncol = 1) +
  geom_boxplot(width = 0.3,
               outlier.shape = "") +
  geom_point(aes(fill = AnyDM,
                 group = sample_names),
             size = 2,
             shape = 21,
             alpha = 0.7,
             position = position_dodge(0.3)) +
  scale_x_discrete("COVID",
                   breaks = c("No_COVID",
                              "COVID"),
                   labels = c("No",
                              "Yes")) +
  scale_y_continuous("Number of hits",
                     breaks = seq(from = 1,
                                  to = 20,
                                  by = 2),
                     labels = (2^seq(from = 1,
                                     to = 20,
                                     by = 2)) - 1) +
  scale_fill_discrete("Any DM",
                      breaks = c(0, 1),
                      labels = c("No",
                                 "Yes")) +
  theme_bw() +
  theme(legend.position = "bottom")

p5 <- ggplot(dt_f3,
             aes(x = AnyDM,
                 y = log2(Hits + 1))) +
  facet_wrap(~ gene_name,
             scale = "free_y",
             ncol = 1) +
  geom_boxplot(width = 0.3,
               outlier.shape = "") +
  geom_point(aes(fill = Cohort,
                 group = sample_names),
             size = 2,
             shape = 21,
             alpha = 0.7,
             position = position_dodge(0.3)) +
  scale_x_discrete("Any DM",
                   breaks = c(0, 1),
                   labels = c("No",
                              "Yes")) +
  scale_y_continuous("Number of hits",
                     breaks = seq(from = 1,
                                  to = 20,
                                  by = 2),
                     labels = (2^seq(from = 1,
                                     to = 20,
                                     by = 2)) - 1) +
  scale_fill_discrete("COVID",
                      breaks = c("No_COVID",
                                 "COVID"),
                      labels = c("No",
                                 "Yes")) +
  theme_bw() +
  theme(legend.position = "bottom")

tiff(filename = "tmp/fig3_covid_dm.tiff",
     height = 10,
     width = 6,
     units = 'in',
     res = 600,
     compression = "lzw+p")
grid.arrange(p4,
             p5,
             nrow = 1)
graphics.off()

grid.arrange(p4,
             p5,
             nrow = 1)
```

### DESeq: AnyDM/no DM in COVID patients only
```{r}
samples_covid_only <- rnaseq_meta_clin$sample_names[rnaseq_meta_clin$Cohort == "COVID"]
ndx_covid_only <- which(colnames(dt_rna_cod) %in% samples_covid_only)

tmp <- as.matrix(dt_rna_cod[, ndx_covid_only, with = FALSE])
rownames(tmp) <- dt_rna_cod$gene_id

dds_cod_dm_covid_only <- DESeqDataSetFromMatrix(countData = tmp, 
                                                colData = rnaseq_meta_clin[sample_names %in% samples_covid_only],
                                                ~ AnyDM)
```

#### Filter out genes with small number of hits
```{r}
keep <- rowSums(counts(dds_cod_dm_covid_only) >= 10) >= 10
sum(keep)

dds_cod_dm_covid_only <- dds_cod_dm_covid_only[keep, ]
```

#### Run DESeq
```{r}
deseq_cod_dm_covid_only <- DESeq(dds_cod_dm_covid_only)

resultsNames(deseq_cod_dm_covid_only)
```

#### DESeq results
```{r}
res_dm_covid_only <- results(deseq_cod_dm_covid_only)
res_dm_covid_only[1:10, ]
rownames(res_dm_covid_only)[1:10]
```


```{r}
out_dm_covid_only <- data.table(gene_id = rownames(res_dm_covid_only),
                                baseMean = res_dm_covid_only$baseMean,
                                log2FoldChange = res_dm_covid_only$log2FoldChange,
                                log2FoldChangeSE = res_dm_covid_only$lfcSE,
                                stat = res_dm_covid_only$stat,
                                pvalue = res_dm_covid_only$pvalue,
                                padj = res_dm_covid_only$padj)

out_dm_covid_only <- merge(dt_rna_cod[, c("gene_name",
                                          "gene_id")],
                           out_dm_covid_only,
                           by = "gene_id")

out_dm_covid_only <- out_dm_covid_only[!is.na(padj), ]

setorder(out_dm_covid_only, 
         padj)


head(out_dm_covid_only)

write.csv(out_dm_covid_only,
          file = "tmp/deseq2_cod_dm_covid_only.CSV",
          row.names = FALSE)
```

```{r}
tmp7 <- out_dm_covid_only[abs(log2FoldChange) > 1 &
                            padj < 0.05, ]
```

#### Calculate the mean expressions of the selected genes
```{r}
tmp <- dt_rna_cod[gene_name %in% tmp7$gene_name, ]
tmp <- melt.data.table(data = tmp,
                       id.vars = 1,
                       measure.vars = 11:ncol(tmp),
                       variable.name = "sample_names",
                       value.name = "Hits")

tmp <- merge(tmp,
             rnaseq_meta_clin,
             by = "sample_names")
```

#### Table 3: DM/no DM significant DEs
```{r}
# med <- tmp[, .(med = median(Hits)),
#            by = list(gene_name,
#                      Cohort)]
# med <- dcast.data.table(data = med,
#                         gene_name ~ Cohort,
#                         value.var = "med")
# med
# 
# t4 <- data.table(gene_name = tmp6$gene_name,
#                  `Log2 Fold Change` = round(tmp6$log2FoldChange, 2),
#                  `SE of Log2 Fold Change` = round(tmp6$log2FoldChangeSE, 2),
#                  `Adjusted p-Value` = round(tmp6$padj, 3))
# 
# t4 <- merge(med,
#             t4,
#             by = "gene_name")
# 
# names(t4)[1:3] <- c("Gene",
#                     "Median hits in non-COVID",
#                     "Median hits in COVID")
# 
# write.csv(t4,
#           file = "tmp/t4.csv",
#           row.names = FALSE)
# t4
```

```{r}
# sum(t4$`Log2 Fold Change` > 0)
# sum(t4$`Log2 Fold Change` < 0)
```

# Session
```{r}
sessionInfo()
```